Almost orthogonal vectors

### 2014-12-09 project3/paper/paper.tex
added the lower bound section

The accuracy of our method can be noted by looking at the computed values for $n=2$ -- the computed values for different values of $N$ are very close to the expected value of $\cos^2 \pi/N$, that we already proved in Section $2$, which makes us confident about the accuracy of this numeric method.
\section{Applications to sphere packing}
We should be able to prove something about the radius of spheres you can pack in the case of $N=n+1$, taking as intuition square and hexagonal tilings of the plane and cubic and tetrahedral tilings of 3-space.

>>>>

The accuracy of our method can be noted by looking at the computed values for $n=2$ -- the computed values for different values of $N$ are very close to the expected value of $\cos^2 \pi/N$, that we already proved in Section $2$, which makes us confident about the accuracy of this numeric method.

\section{A lower bound}

We can use a well-known lemma to provide a useful lower bound on $\epsilon$. Our presentation of the result follows that of a paper by Noga Alon\footnote{Lemma 2.2, \emph{Perturbed identity matrices have high rank: proof and applications}, Noga Alon.}.

\begin{lemma}
Let $A=(a_{i,j})$ be an $n$-by-$n$ real symmetric matrix with $a_{i,i}=1$ for all $i$ and $|a_{i,j}|\leq \sqrt{\epsilon}$ for all $i\neq j$. Let $d$ be the rank of $A$. Then
\[\epsilon\geq \frac{n-d}{d(n-1)}.\]
\end{lemma}
\begin{proof}
We define $\lambda_1,\ldots, \lambda_n$ to be the eigenvalues of $A$; write $k$ as the number of nonzero eigenvalues. Their sum is the trace of $A$, equal to $n$. Recall that the number of nonzero eigenvalues of a complex matrix is less than or equal to the rank of the matrix: so we have that $d\geq k$.

Recall the Cauchy-Schwarz for $\mathbb R^n$: given $(x_1,\ldots, x_n)$ and $(y_1,\ldots, y_n)\in \mathbb R^n$, we have
\[\left(\sum_{i=1}^nx_iy_i\right)^2\leq \left(\sum_{i=1}^n x_i^2\right)\left(\sum_{i=1}^ny_i^2\right).\]
In our case, let the $x_i$ equal $\lambda_i$, and let each $y_i$ equal 1 when $\lambda_i$ is nonzero and zero when $\lambda_i$ is zero. We get
\[n^2=\left(\sum_{i=1}^n \lambda_i\right)^2\leq \left(\sum_{i=1}^n \lambda_i^2\right) k\]
so that
\[\sum_{i=1}^n \lambda_i^2\geq\frac{n^2}{k}\geq \frac{n^2}{d}.\]
In fact, the sum $\sum_{i=1}^n \lambda_i^2$ is equal to the trace of $A^TA$, which we can compute explicitly to be $\sum_{i,j}a_{i,j}^2$. Hence we have that
\[\sum_{i,j}a_{i,j}^2\geq \frac{n^2}{d}.\]
But because $|a_{i,j}|\leq \sqrt\epsilon$ for $i\neq j$, we can bound the left hand side:
\[n+n(n-1)\epsilon\geq \left(\sum_{i=1}^j a_{i,i}^2\right)+\left(\sum_{i\neq j}a_{i,j}^2\right)=\sum_{i,j}a_{i,j}^2.\]
We obtain, therefore, that
\[n+n(n-1)\epsilon\geq \frac{n^2}{d},\]
and it clearly follows that
\[\epsilon\geq \frac{n-d}{d(n-1)}.\]
\end{proof}
We can now prove our lower bound. 
\begin{theorem}
\label{lower_bound}
For any choice of $n$ and $N$, the minimum value of $\epsilon$ satisfies
\[\epsilon\geq \frac{N-n}{n(N-1)}.\]
\end{theorem}
\begin{proof}
Pick any collection of $N$ unit vectors $v_1,\ldots, v_N$ in $\mathbb R^n$, and define $\epsilon$ as usual. Let us define the $n$-by-$N$ matrix
\[A=\left(\begin{matrix}
| & | & \cdots & | & |\\
v_1 & v_2 & \cdots & v_{N-1} & v_N\\
| & | & \cdots & | & | \end{matrix}\right).\] 
Now consider the $N$-by-$N$ Gram matrix:
\[G(v_1,\ldots, v_N)=A^TA=\left(\begin{matrix}
v_1\cdot v_1 & v_1\cdot v_2 & \ldots & v_1\cdot v_n\\
v_2\cdot v_2 & v_2\cdot v_2 & \ldots & v_2\cdot v_n\\
\vdots & \vdots & \ddots & \vdots\\
v_n\cdot v_1 & v_n\cdot v_2 & \ldots & v_n\cdot v_n\end{matrix}\right).\]
Let $d$ be the rank of $G$; since $A$ has rank at most $n$, we have that $d\leq n$. Note that $G$ is real and symmetric, with $1$s along the diagonal and all other entries with absolute value bounded by $\sqrt \epsilon$. We can apply the lemma to conclude
\[\epsilon\geq \frac{N-d}{d(N-1)}.\]
Since $d\leq n$, we have as desired
\[\epsilon\geq \frac{N-n}{n(N-1)}.\]
\end{proof}
For $N=n+1$, for instance, the theorem provides us a bound of 
\[\epsilon\geq \frac{(n+1)-n}{n(n+1-1)}=\frac 1{n^2}\]
implying that the $n$-simplex is indeed an optimal configuration. Similarly, for $n=3, N=6$, the theorem gives us
\[\epsilon\geq \frac{6-3}{3(6-1)}=\frac{1}{5}.\]
Again, the configuration we obtained was optimal. We present now a table of the bounds the theorem gives to us for varying $n$ and $N$:
\begin{table}[h]
   \centering
    \begin{tabular}{ | c | c | c | c |}
    \hline
      & $n=2$ & $n=3$ & $n=4$ \\ \hline
    $N=2$ & $0$ & $-$ & $-$ \\ \hline
    $N=3$ & $1/4$ & $0$ & $-$ \\ \hline
    $N=4$ & $1/3$ & $1/9$ & $0$ \\ \hline
    $N=5$ & $3/8$ & $1/6$ & $1/16$ \\ \hline
    $N=6$ & $2/5$ & $1/5$ & $1/10$ \\ \hline
    $N=8$ &  $3/7$ & $5/21$ & $1/7$ \\ \hline
    $N=12$ & $5/11$ & $3/11$ & $2/11$ \\
    \hline
    \end{tabular}
    \caption {Lower bounds on $\epsilon$ for different values of $(n, N)$, given by Theorem \ref{lower_bound}.}
    \end{table}

Note that the bounds for $n=2$ are not tight, as we proved in Section 2 that the minimum $\epsilon$ was given by $\cos^2(\pi/N)$. One direction for further research would be checking whether the bounds provided by Theorem \ref{lower_bound} are indeed tight for other combinations of $n$ and $N$.

\section{Applications to sphere packing}
We should be able to prove something about the radius of spheres you can pack in the case of $N=n+1$, taking as intuition square and hexagonal tilings of the plane and cubic and tetrahedral tilings of 3-space.
